A primer to numerical simulations: The perihelion motion of Mercury

### 2017-12-01 tex/MercuryNumerical.tex
Added appendix

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\python}[0]{\texttt{Python} }
\newcommand{\abs}[1]{\left\vert #1 \right\vert}

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{url once uploaded (arXiv)}.

@@

In \python, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 10
\end{lstlisting}
Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $dt$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
	#<...Code...>
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
\end{lstlisting}
\python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
Furthermore, \python is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the finalization of the function.

@@

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to include further objects of the \texttt{VPython} module \texttt{visual}.
Additional to the vector class, one has to include
\begin{lstlisting}
from vpython import vector, sphere, color, curve, rate
\end{lstlisting}
The class sphere will represent mercury and the sun in the simulation
\begin{lstlisting}
# define the initial coordinates; M = mercury, S = sun
M = sphere( pos=vec_rM0,       radius=0.5,  color=color.red   )
S = sphere( pos=vector(0,0,0), radius=1.5,  color=color.yellow)
# and the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
# add a visible trajectory to mercury
M.trajectory = curve(color=color.white)
\end{lstlisting}
We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.

Last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
\begin{lstlisting}
t     = 0
alpha = 0
# update the objects
while t < T:
	# set the frame rate (you can choose a higher rate to accelerate the program)
	rate(100)
	# update the drawn trajectory with the current position
	M.trajectory.append(pos=M.pos)
	# update the velocity and position
	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
	# update time
	t = t + dt
\end{lstlisting}

@@

There are several ways to find the position of the perihelion.
The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
This can be done within the $while$-loop.
To find identify a minimal distance to the sun, one has to know the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} in addition to the current position.
If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
The implementation can be realized as follows
\begin{lstlisting}
# set up vectors
vec_r_last = vec_rM0
turns      = 0
max_turns  = 10
list_perih = list()
\end{lstlisting}
\begin{lstlisting}
# find perihelion for each turn and print it out
while turns < max_turns:
	vec_r_before_last = vec_r_last
	# store position of mercury in a new vector (since we will change M.pos)
	vec_r_last = vector(M.pos)
	#<...update mercury position...>
	# check if at perihelion
	if (vec_r_last.mag < M.pos.mag) and (vec_r_last.mag < vec_r_before_last.mag):
		list_perih.append(vec_r_last)
		turns = turns+1
\end{lstlisting}
To compute the angle between two vectors one uses the following
formula
 \begin{equation}
 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
	\, .
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# import function to resolve angle
from vpython import acos, pi, dot
# define function for angle extraction
def angle_between(v1, v2):
	return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
\end{lstlisting}
To account for the statistical errors (e.g., numerical rounding errors) one can be average over a few turns.
Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
\begin{lstlisting}
sum_angle=0.
for n in range(1, max_turns):
	# calculate angle
	sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
# display the average
print(sum_angle/(max_turns-1.))
\end{lstlisting}

As explained above [and further discussed in chapter \ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$\footnote{%
	Of course for very large values of $\alpha$ this linear relation will eventually break down.
	We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.
}.
The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
This can be done either by hand, \python (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Instructions to install VPython}\label{appendix:python}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{@CK: installation python+vpython win+linux}
\subsection{Mac}

\href{http://vpython.org/contents/download_mac.html}{Download VPython for Mac}

%------------------------------------------
\subsection{Running the Program}
%------------------------------------------
\todo[inline]{@CK: running python win+linux}
The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
Depending on the editor of choice, there might already be an execute functionality.
In case one uses a linux like operating system and can execute the code from a console/terminal by running the command
\begin{center}
	\texttt{python 'address/to/python/file.py'}
\end{center}
If one already works in the file directory and has used the suggested naming scheme, this reads 
\begin{center}
	\texttt{python mercury\_orbit.py}
\end{center}
Here, the specified \texttt{python} interpreter needs to agree with the installed \texttt{VPython} version.
For more information see \ref{appendix:python}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The code}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo[inline]{@CK: inline code + script + test code + confirm data}
Here we should put the code

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Commands
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\python}[0]{\texttt{Python}}
\newcommand{\vpython}[0]{\texttt{VPython}}
\newcommand{\abs}[1]{\left\vert #1 \right\vert}
\newcommand{\code}[1]{{\scriptsize\colorbox{light-gray}{\texttt{#1}}}}

@@

In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
particular
the perihelion motion of Mercury.  We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.

@@

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.e-7  # in units of R0
\end{lstlisting}
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 10
\end{lstlisting}
Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracys.
	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $dt$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
\begin{lstlisting}
# Define the coordinate and velocity update function
def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
	#<...Code...>
	return vec_rM_new, vec_vM_new

# Call the function
vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
\end{lstlisting}
\python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
Furthermore, \python{} is an Interpreter language.
Each line of the code is executed when the Interpreter passes it.
For this reason, all the variables defined before the function are global and thus known to the function.
The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the finalization of the function.

@@

%------------------------------------------
\subsection{Visualizing the Motion with VPython}
%------------------------------------------
To start the visualization one has to include further objects of the \texttt{VPython} module \texttt{visual}.
Additional to the vector class, one has to include
\begin{lstlisting}
from vpython import vector, sphere, color, curve, rate
\end{lstlisting}
The class sphere will represent mercury and the sun in the simulation
\begin{lstlisting}
# define the initial coordinates; M = mercury, S = sun
M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
# and the initial velocities
M.velocity = vec_vM0
S.velocity = vector(0,0,0)
# add a visible trajectory to mercury
M.trajectory = curve(color=color.white)
\end{lstlisting}
We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.

Last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
\begin{lstlisting}
t     = 0
alpha = 0
# update the objects
while t < 2*TM:
	# set the frame rate (you can choose a higher rate to accelerate the program)
	rate(100)
	# update the drawn trajectory with the current position
	M.trajectory.append(pos=M.pos)
	# update the velocity and position
	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
	# update time
	t = t + dt
\end{lstlisting}

@@

There are several ways to find the position of the perihelion.
The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
This can be done within the $while$-loop.
To find identify a minimal distance to the sun, one has to know the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} in addition to the current position.
If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
The implementation can be realized as follows
\begin{lstlisting}
# set up vectors
vec_r_last = vec_rM0
turns      = 0
max_turns  = 10
list_perih = list()
# find perihelion for each turn and print it out
while turns < max_turns:
	vec_r_before_last = vec_r_last
	# store position of mercury in a new vector (since we will change M.pos)
	vec_r_last = vector(M.pos)
	#<...update mercury position...>
	# check if at perihelion
	if (vec_r_last.mag < M.pos.mag) and (vec_r_last.mag < vec_r_before_last.mag):
		list_perih.append(vec_r_last)
		turns = turns+1
\end{lstlisting}
To compute the angle between two vectors one uses the following
formula
 \begin{equation}
 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
	\, .
 \end{equation}
This is readily implemented in \texttt{VPython} via
\begin{lstlisting}
# import function to resolve angle
from vpython import acos, pi, dot
# define function for angle extraction
def angle_between(v1, v2):
	return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
\end{lstlisting}
To account for the statistical errors (e.g., numerical rounding errors) one can be average over a few turns.
Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
\begin{lstlisting}
sum_angle=0.
for n in range(1, max_turns):
	# calculate angle
	sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
# display the average
print(sum_angle/(max_turns-1.))
\end{lstlisting}

As explained above [and further discussed in chapter \ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$\footnote{%
	Of course for very large values of $\alpha$ this linear relation will eventually break down.
	We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.
}.
The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Instructions to install VPython}\label{appendix:python}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In short, to run the above commands and scripts, one needs to install \python{} and \vpython{}.
We recommend the newest \vpython{} version 7 or higher which best runs with the \python{} versions of 3.5 or higher.
It is also possible to use \python{} 2.7 or higher, but the 3.x versions have better \vpython{} support.

For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} [see also \url{http://vpython.org}].

Below, we provide detailed descriptions how to install these versions of \python{} and \vpython{} for different operating systems.
Independent of the operating systems, one has to execute the following steps:
\begin{enumerate}
	\item install and update \python{} (at best 3.x, x$\geq5$),
	\item install and update a \python{} package manager,
	\item install \vpython{} and dependencies.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Linux based systems (including macOS)}\label{appendix:python-max}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
All Linux based system come with an installation of \python{}.
Thus, if not already at the newest version, one just have to update existing installations.
This is best done by using a distribution manager.
Native to Linux based systems is for example \texttt{dpkg} and \texttt{apt-get}, or \texttt{homebrew} and \texttt{MacPorts} for macOS.
While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
For this we refer to \url{https://www.macports.org/install.php} or \url{https://docs.brew.sh/Installation.html}.
At first one should open a terminal/console (bash).
Depending on the installation, one usually needs superuser rights in order to update \python{}.
Therefore one has to prepend \code{sudo} to the following commands and confirm them by entering the superuser password.
The installations using a distribution manager are save and tested by the vast community.
In the following, the successive order of commands to update python is presented for \texttt{apt-get} (Linux) and \texttt{MacPorts} (macOS).
\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{l || l | l}
	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
	1. Update the distribution manager & \code{sudo apt-get update} & \code{sudo port selfupdate} \\
	2. Install \& update \python{} & \code{sudo apt-get install python3} & \code{sudo port install python36} \\
	3. Installing \python{} package manager & \code{sudo apt-get install python3-pip} & \code{sudo port install py36-pip}
\end{tabular}
\normalsize
\end{table}

Depending on the previously installed packages, this command might install further dependencies automatically (after confirmation).
One should test if the installation was successful by typing \code{python3 --version} or \code{python3.6 --version} (depending on the installation name) which prints out the version of the python installation (and should be larger than 3.5).
Next one must update the package manager to get the newest versions of python packages.
This is done by running
\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{l || l | l}
	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
	4. Update \code{pip} & \code{pip3 install --upgrade --user pip wheel} & \code{pip-3.6 install --upgrade --user pip wheel} \\
\end{tabular}
\normalsize
\end{table}

Note that this just upgrades the packages for the current user.
If one wants the systemwide installation, one again has to prepend \code{sudo} and can ignore the \code{--user} flag (\code{sudo pip3 install --upgrade pip wheel}).
Furthermore, depending on the installed version of \code{pip}, it could have a different name.
This is best figured out by typing \code{pip} and pressing \texttt{tab}.
Finally, one can install \vpython{} by running
\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{l || l | l}
	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
	5. Install \vpython{} & \code{pip3 install --user vpython} & \code{pip-3.6 install --user vpython} \\
\end{tabular}
\normalsize
\end{table}

At this point, everything needed is installed and one can start the simulation.
To test the \vpython{} installation and should run \code{python3} or \code{python3.6}, which starts a \python{} console.
Next, one shall run \code{from vpython import *}, which loads the \vpython{} package and opens the default browser.
Running \code{box()} should generate a white box in the browser window, which can be rotated using the mouse.
The program is quit by running \code{quit()} or pressing \texttt{Ctrl-c}.

In general, there are multiple ways to run python scripts.
For the sake of this paper, we suggest two scenarios.

The first scenario to run python scripts is that one writes and saves a \python{} file, e.g., \texttt{test.py}, which can be edited using any text editor of choice.
To run the python script one must navigate the terminal to the containing folder using the \code{cd} (and \code{ls}) command.
While \code{ls} prints out the folders and files in the current directory, \code{cd} changes the directory.
The command \code{cd} either takes relative paths (directories which are in the current directory), e.g., \code{cd folder}, or absolute paths beginning with a \code{/}, e.g., \code{cd /Users/name/Desktop/}.
Once one has navigated to folder which contains the python file --- e.g., the file name is listed when typing \code{ls} --- than one can type \code{python3 test.py} or \code{python3.6 test.py} to execute the script.
Again, this script can be ended (if it does not end automatically) by pressing \texttt{Ctrl-c}.

The second scenario to run \python{} scripts is given by the interactive \texttt{ipython} or \texttt{jupyter} notebooks.
This feature is automatically installed when installing the newest version of \vpython{}.
To start an interactive session, one has to run \code{jupyter notebook} (or equivalent, depending on the installation name. Again, this can be figured out by typing \code{jupyter} and pressing \texttt{tab}).
This will print a link (containing the name \texttt{localhost}) and open the browser at the link-address.
Within this browser window, one can freely navigate to the directory of choice and create interactive \python{} and \vpython{} sessions by clicking on existing files or creating new ones (upper right button \texttt{new}$\rightarrow$\vpython{}).

We provide example files for both scenarios.\todo{@arXiv: online link}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Windows}\label{appendix:python-windows}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
Following the link \url{https://conda.io/docs/user-guide/install/download.html}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
Both are equally sufficient, because one can install further updates after the initial installation.
In the case of \texttt{Miniconda} (\url{https://conda.io/miniconda.html}) one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.

Installing one of the two executables creates a new windows application called \texttt{Anaconda Prompt}.
Starting this application opens a terminal which we will user to install the required \python{} packages.
At first, one should test if the \texttt{Anaconda} installation was successful by running the command \code{conda list}.
This command should print a list of already installed packages.
Also one should run \code{conda update conda} to make sure one has the most recent installation.

To install \vpython{} one simply has to run \code{conda install -c vpython vpython}.
The additional flag \code{-c vpython} tells \texttt{Anaconda} the "channel" where to look for the \vpython{} package.
Running this command prints out a list of dependencies which must be installed in order for \vpython{} to work --- this must be confirmed.

In principle this is already sufficient to run \vpython{} scripts.
One could furthermore install \texttt{Anaconda}s native navigator \texttt{anaconda-navigator} by running \code{conda install anaconda-navigator}.
This will install another application, \texttt{Anaconda Navigator}, which can be run from the Desktop and makes it easier to program in \python{}.

Again there are several possibilities to run \python{} files on \texttt{Windows}.
\begin{itemize}
	\item One could create a file \texttt{test.py}, edit it with the editor of choice and specify that this file should be opened with \texttt{Python3}.
		To do so, \texttt{right click} on  the file, select \texttt{open with} and locate the \texttt{Anaconda Python3} installation.
		Programs can be closed by closing the running window.
	\item Start the \texttt{Anaconda Navigator} application, open an interactive \texttt{qtconsole} and program within the newly opened terminal.
		Programs can be closed by closing the running window.
	\item Start the \texttt{Anaconda Navigator} application and click on the \texttt{jupyter notebook} button.
		This opens an interactive \python{} session in the default web-browser.
		Here one open existing notebooks or create new files (upper right button \texttt{new}$\rightarrow$\vpython{}).
\end{itemize}

To start and test the \vpython{} installation one should pick any out of the three options and execute the code
\begin{lstlisting}
from vpython import *
box()
\end{lstlisting}
Running this code should open the default browser and render a white box on black background.
Using the mouse (right click and drag), one should be able to rotate the box.
It was found that the \texttt{Internet Explorer} is not able to properly render the animation.
Here we suggest to use the \texttt{Firefox} web-browser as the default choice.

Once this test case works, one can start to test the mercury orbit simulation.
We provide example files for both \texttt{juypter notebooks} and regular \python{} files.

\todo{@arXiv: online link}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
\section{The code}\label{sec:code}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Below we provide an example script on how to program the visualization of the mercury orbit in \vpython{}.
\lstinputlisting[language=Python]{mercury_orbit_base.py}
Further examples and template files can be found online\cite{}.\todo[inline]{@arXiv: online link}
